---
code-annotations: hover
citation-location: margin
---

# Bivariate Regression with Binary & Categorical Predictors {#sec-binary-predictors}

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(rio)
library(tidyverse)

demdata <- import("data/demdata.rds") |> 
  as_tibble() 
```

Our last chapter concluded by showing how to perform a linear regression with a continuous (interval/ratio) independent variable. In this chapter we'll see how to use a linear regression model with a binary or categorical independent variable.

As always, we begin our R script by loading relevant libraries and by loading our data. Note that these libraries are already installed on university computers but must be loaded prior to use.

```{r}
#| eval: false

#Packages
library(rio)          #loading data
library(tidyverse)    #data manipulation and plotting

#Import our data
demdata <- import("demdata.rds") |> 
  as_tibble()  # <1>
```

1.  You do not always need `as_tibble()` as shown here. We do this here because it is easier to show larger datasets with them. See [Statistics I, 2.1](https://poweleiden.github.io/statistics1/data_02_filtering_selecting.html#tibbles){target="_blank"}.

## Data Preparation: Converting to factor variable

One advantage of linear regression models is that we can predict a dependent variable with different *types* of predictor variable, including *binary* and *categorical* independent variables. In doing so, our regression will compare the (predicted) average value of the dependent variable in one category against the (predicted) average in the other category or categories.

In order to use binary and categorical variables as predictors in a regression model we need to include them as a dichotomous or "[dummy](https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faqwhat-is-dummy-coding/){target="_blank"}" (0/1) variable. One dummy variable is used when the variable is binary while multiple dummies are included when the variable is categorical.[^linear_02-1] R automatically creates dummies for factor variables used as predictors in regression models, so we will always convert our binary/categorical variables to factor variables before including them in a regression.[^linear_02-2]

[^linear_02-1]: Specifically, we include k-1 dummies, where k = the number of categories. If the categorical variable has four categories (for instance: North, West, South, and East), then we include three dummy variables in the model. If it has two categories (i.e., a binary variable), then we include one dummy variable in the model.

[^linear_02-2]: In some circumstances the variable may already be stored as factor variable in our dataset enabling us to skip this first step. However, we may need to change what category is used as the reference category. See @sec-relevelling.

Here we will convert a binary variable named `TYPEDEMO1984`, and a categorical variable named `Typeregime2006`, into factor variables using the procedures described in @sec-factor-command. Note that we create entirely new variables when doing so.

```{r}
# Convert to factor
demdata <- demdata |> 
  mutate(
    # Binary example
    TYPEDEMO1984_factor = factor(TYPEDEMO1984, 
                        levels = c(1, 2), 
                        labels = c("Autocracies", "Democracies")), 
    # Categorical example
    Typeregime2006_factor = factor(Typeregime2006, 
                            levels = c(1, 2, 3), 
                            labels = c("Liberal democracy", 
                                       "Electoral democracy", 
                                       "Autocracy")))

# Check your work!
levels(demdata$TYPEDEMO1984_factor)
table(demdata$TYPEDEMO1984_factor, useNA = "always")
table(demdata$TYPEDEMO1984_factor, demdata$TYPEDEMO1984, useNA = "always")

levels(demdata$Typeregime2006_factor)
table(demdata$Typeregime2006_factor, useNA = "always")
table(demdata$Typeregime2006_factor, demdata$Typeregime2006, useNA = "always")

```

::: callout-important
## Warning!

We did two important things here: (1) we did not overwrite the initial variable when creating the factor and (2) we double checked our work to make sure that the factor creation process led to the expected output. This is particularly important with factor variables because a failure to properly set them up can lead to errors when performing a linear regression model. For instance, mistakenly creating a factor where only one level of the variable has non-NA values will prevent a regression model from running as we discuss in @sec-contrasts-can-be-applied-only-to-factors-with-2-or-more-levels.
:::

## Including in a Model and Interpreting Coefficients

We include binary/categorical variables in a regression model in the same way that we did for a continuous variable. For instance:

```{r}
# Using the binary variable as a predictor: 
model_binary <- lm(v2x_polyarchy ~ TYPEDEMO1984_factor, data=demdata)
summary(model_binary)

# Using the categorical variable as a predictor: 
model_categorical <- lm(v2x_polyarchy ~ Typeregime2006_factor, data=demdata)
summary(model_categorical)
```

::: callout-note
#### Output Explanation

The output from a model with a binary/categorical variable is the same as when the predictor is continuous with one exception: R will format the variable names differently in the Coefficients area depending on whether the variable is a factor variable or not. If the variable is not a factor variable, then R will simply show the variable name (e.g., "gini_2019"). If the variable is a factor, then the display is formatted as so: "variableCategory". For instance: "TYPEDEMO1984_factorDemocracies" or "regime2006_fctAutocracy".
:::

There are some subtle differences in how we interpret the coefficients of a model that (only) includes a factor variable as a predictor variable:

::: callout-warning
#### Interpretation

The `Estimate` column provides the coefficient values from our regression model.

The "(Intercept)" row gives you the coefficient for the Intercept: What is the average value of the DV we expect to observe based on this model if all of the included IVs = 0. If the only predictor variable in the model is a factor variable, then the (Intercept) value = the mean of the dependent variable among observations in the initial level of the factor variable, what we often call the "reference group".

Here, for instance, is the mean value of `v2x_polyarchy` among countries with a value of "Liberal Democracy" on the `TYPEDEMO84_factor;` it is identical to the (Intercept) value above.

```{r}
demdata |> 
  filter(TYPEDEMO1984_factor == "Autocracies") |> # <1> 
  summarize(mean_democracy = mean(v2x_polyarchy, na.rm=T)) |> 
  as.data.frame() # <2>

```

1.  This removes all observations that do not have the value of "Autocracies" on our factor variable
2.  This option is used to force R to show you all the digits of the mean value so you can compare it to the Intercept value reported in the summary output

The coefficient for a continuous predictor variable is interpreted as the slope of a line (e.g., how much does Y change, on average, with each one unit change in X?). The coefficient(s) for a factor variable, on the other hand, are best discussed as telling us the difference in the mean value of the DV between the category named in the output and the reference group (the first level of the factor). The coefficient for "TYPEDEMO1984_factorDemocracies", for instance, is `r coef(model_binary)[2]`.[^linear_02-3] We would interpret this as telling us that the average value of the dependent variable among countries with a "Democracies" values on `TYPEDEMO1984_factor` is `r coef(model_binary)[2]` scale points greater than the average value among countries with a value of "Autocracies' level (our reference group) on that variable.

We can again see this by looking at the underlying average values:

```{r}
# The averages
demdata |> 
  group_by(TYPEDEMO1984_factor) |> 
  summarize(mean_democracy = mean(v2x_polyarchy, na.rm = T)) |>
  as.data.frame() 

# Avg. in Democracies - Avg. in Autocracies
 0.6902456 - 0.4175698

```

The same goes for models with categorical factor variables. The "(Intercept)" value in `model_categorical` above tells us the average value of the DV among observations in the reference group of the factor variable (there: "Liberal Democracy") while the coefficients for the factor variable tell us the difference from this average score. The average 2020 democracy score among countries that are coded as an "Electoral Democracy" in 2006 is `r round(coef(model_categorical)[2],2)` scale points lower than the average 2020 democracy score among the "Liberal Democracy" reference group countries, for instance.

```{r}
demdata |> 
  group_by(Typeregime2006_factor) |> 
  summarize(mean_democracy = mean(v2x_polyarchy, na.rm=T)) |> 
  as.data.frame() 

# Avg. in Elec Democracy - Avg. in Lib Democracy
0.4329811 - 0.7540423

# Avg. in Autocracy - Avg. in Lib Democracy
0.2482683 - 0.7540423
```

See @sec-reporting-linear-regression for how to report these values in our assignments and in formal papers.
:::

[^linear_02-3]: We would normally round this to 2 or 3 digits, but we show you the whole coefficient here so you can see how it compares to the difference in means.
