---
code-annotations: hover
---

# Logistic Regression & Odds Ratios {#sec-logistic-regression-odds-ratios}

Logistic regression models are used to model binary outcome variables. We will use survey data from the Netherlands which was collected as part of Round 9 of the [European Social Survey](https://www.europeansocialsurvey.org){target="_blank"} for our examples. The dataset is available in SPSS format (`.sav`) from the ESS website.

We begin, as always, by loading the libraries that we will use in our analyses as well as our data.

```{r}
#| eval: false

#Packages
library(sjPlot)          #checking variable names, values, and labels
library(broom)           #for obtaining summaries of regression models
library(rio)             #loading data
library(tidyverse)       #data manipulation and plotting

#Data
ESS9NL <- import("ESS9e03, Netherlands.sav")

```

```{r}
#| echo: false
#| message: false
#| warning: false

#Packages
library(sjPlot)            #checking variable names, values, and labels
library(broom)             #Model summaries, including coefficients
library(rio)               #Importing data
library(tidyverse)         #Data management & plotting

#Data
ESS9NL <- import("data/ESS9e03, Netherlands.sav")

```

## Performing a Logistic Regression

### Data Preparation

Our example will investigate the relationship between self-reported gender (`gndr`) and voter turnout (i.e., did the person report voting in an election or not; `vote`). Our goal here is to predict whether a person reported voting in the most recent election.

The dependent variable in a logistic regression model needs to be either a 0/1 numeric variable or a factor variable where the non-baseline (or upper) level represents the outcome that we are trying to predict with our model. Independent variables, meanwhile, need to be properly handled as well (e.g., handling missing data and converting binary/categorical predictors to factor variables). Per @sec-investigating-datasets, we can use the `view_df()` from from the `sjPlot` library to inspect the two variables of interest to us (`gndr` and `vote`). We can also use the `table()` function to get a further sense of the characteristics of these two variables given that neither is continuous in nature.

```{r}
#Variable attributes
ESS9NL |> 
  select(gndr, vote) |> 
  view_df()

#Tabulation
table(ESS9NL$gndr, useNA = "always")
table(ESS9NL$vote, useNA = "always")
```

There exist value labels in our dataset associated with missing data codes (e.g., 7 = Refusal, 8 = Don't Know, etc.). However, the outputs of `table()` show that missing data is either already set to NA, or there are no observations in these categories, so we do not need to take any explicit step to handle missing data in this instance.[^logit_01-1]

[^logit_01-1]: In this case, there are no observations with NA values on the `gndr` variable and 5 with NA values on the `vote` variable.

Our predictor variable only has two categories, so we will convert it into a factor variable. Our DV, meanwhile, has there categories with observations in them: Yes, No, and Not Eligible. Per above, we need to convert this into a binary variable prior to analysis.[^logit_01-2] We are not interested in the "Not Eligible" category here (i.e., we want to examine whether a person *chose to vote* vs. *chose not to vote*) so we will convert observations in this category to NA. We can automatically do this by converting `vote` into a factor variable and simply including only the two categories we care about in the `levels = c()` portion of the syntax (see @sec-factor-command). We set "No" (did not vote) as the reference group/baseline level of the factor variable, which will mean that our logistic model will tell us about the probability of saying "Yes" (did vote) rather than "No" (did not vote).

[^logit_01-2]: The logit models that we will use require a binary outcome variable. However, there are logistic models that can be used to analyze categorical (i.e,. 3 or more category) DVs. These are called multinomial logistic model.

```{r}
#Factorize variables
ESS9NL <- ESS9NL |>
  mutate(
    gndr_F = factor(gndr, 
                    levels = c(2, 1), 
                    labels = c("Female", "Male")), # <1>
    vote_F = factor(vote, 
                    levels = c(2, 1), 
                    labels = c("Did not vote", 
                               "Voted"))) # <2>
  

```

1.  This will use observations with a value of 2 as the 'reference' or 'baseline' category in subsequent contrasts (i.e., Male - Female). We could very well flip this around here (Female - Male).
2.  We did not include the value of "3" in the `levels = c()` portion of the syntax. This automatically turns those observations to NA as discussed in @sec-converting-to-a-factor-variable-using-factor .

Let's check our work:

```{r}
# Gender
levels(ESS9NL$gndr_F)
table(ESS9NL$gndr_F, ESS9NL$gndr, useNA = "always")

# Vote
levels(ESS9NL$vote_F)
table(ESS9NL$vote, ESS9NL$vote_F, useNA = "always")

```

Success!

::: callout-important
#### Warning!

The DV in a logistic regression must be a binary variable. Make sure you are creating the factor variable such that the higher level of the variable is the category you are trying to predict membership within. Otherwise, your interpretations might end up being wrong by accident.
:::

### Performing a Logistic Regression

Performing logistic regression in R is very similar to linear regression. However, instead of the `lm()` function, we rely on the `glm()` function, which stands for generalized linear model.

```{r}
#Run the model
Vote_model <- glm(vote_F ~ gndr_F, 
                  data = ESS9NL, 
                  family = "binomial")

```

`Vote_model <-`

:   We assign the results of our estimation to a new object.

`glm(vote_F ~ gndr_F,`

:   We perform `glm` with `vote` as our dependent variable, predicted (`~`) by our only independent variable here: `gndr`. If we want to add more variables, we connect them with a '+' sign.

`data = ESS9NL,`

:   We specify the dataset to be used.

`family = "binomial")`

:   We specify the family for the generalized linear model. For logistic regression this is "binomial". This part of the code remains unchanged. See the Common Errors appendix ( @sec-glm-factor) for an error that could arise if you did not specify the model's family.

Let's take a look at the output using the built in `summary()` command:

```{r}
summary(Vote_model)
```

::: callout-note
#### Output Explanation

The structure of this output is highly similar to what we saw with the output of an `lm()` model.

-   Call: The model being fit
-   Deviance Residuals: This provides some summary data about the model's residuals.
-   Coefficients: This provides the coefficients from the model (Estimate) as well as the coefficient's standard error (Std. Error), a test-statistic (z-value; the Z-Statistic as given by $\frac{\textrm{Coefficient}}{\textrm{Std. Error}}$), and the p-value for the test statistic (Pr(\>\|z\|)). Symbols pertaining to statistical significance may be provided to the right of the p-value with a line indicating how to interpret these symbols provided at the bottom of the Coefficients output ("Signif. Codes:").
-   (Dispersion parameter...): This can be ignored.
-   Area that begins with Null deviance: This area relates to the fit of the model, which will be discussed in a subsequent chapter.
-   Number of Fisher Scoring Iterations: This can be ignored.
:::

We can add multiple predictors to the model in a way similar to the linear regression syntax, by adding them with a `+` symbol. Here add age (`agea`), trust in politicians (`trstplt`), and left-right ideology (`lrscale`). We did not need to take any data management steps with these variables because they are continuous variables and missing data is already coded as NA in our dataset.

```{r}
#Run the model
Vote_model_mp <- glm(vote_F ~ gndr_F + agea + trstplt + lrscale, 
                     data = ESS9NL, family = "binomial")

#Check the output
summary(Vote_model_mp)

```

::: callout-warning
#### Interpretation

Logistic regression coefficients are on the log of the odds scale. The coefficient for `gndr_FMale` in the `vote_model_mp` model tells us the difference in the log of the odds of voting between male and female respondents, while the `agea` coefficient tell us how the log of the odds of voting change with each one unit increase in respondent age.

You can use the direction (positive, negative) and statistical significance of a logistic regression coefficient to talk about the general relationship between the predictor variable and the DV. However, it is not really possible to directly and clearly communicate what a logistic coefficient tells us about the magnitude of the relationship between an IV and the DV because of this log of the odds scaling. You should instead focus on odds ratios (see below), average marginal effects (@sec-marginal-effects), or predicted probabilities (@sec-logit-predicted-probabilities) to give more specific meaning to your discussion.

In this example:

-   Voter turnout is more likely among men than women, but the difference is not statistically significant (p = 0.78) so we cannot rule out the possibility of no difference in voter turnout between the two groups.
-   Voter turnout becomes more likely with age (i.e., older respondents are more likely to vote than younger ones) and this association is statistically significant (p \< 0.001).
-   Voter turnout becomes more likely as trust in politicians increases and this relationship is statistically significant (p \< 0.001).
-   Voter turnout grows more likely as we move from left to right on the ideology scale, but the effect is not statistically significant (p = 0.46) so we cannot rule out the possibility that a one unit change in ideology actually does not lead to any change in the chances of turning out to vote.
:::

## Odds Ratios

Odds ratios are one way that we can translate logistic regression coefficients into something easier to interpret and communicate.

We can use the `tidy()` command from the `broom` package to toggle between output expressed in the log of the odds scale (our logistic coefficients from above) and odds ratios.

```{r}
#tidy with logistic regression and confidence intervals
tidy(Vote_model_mp, conf.int = TRUE)

#tidy with odds ratios and confidence intervals
tidy(Vote_model_mp, conf.int = TRUE, exponentiate = TRUE)

```

Here is how to read the syntax for the latter command:

`tidy(Vote_model_mp`

:   We apply the tidy function on the model specified in brackets.

`conf.int = TRUE`

:   We ask for the confidence intervals for the logistic regression coefficients or odds ratios. We can write 'FALSE' or leave out this statement if confidence intervals are not needed.

`exponentiate = TRUE)`

:   We ask for the exponentiated logistic regression coefficients, which are the odds ratios. We can shorten this to `exp = TRUE` and obtain the same results. We can write 'FALSE' or leave out this statement if we want the logistic regression coefficients.

::: callout-warning
#### Interpretation

There are three important things to keep in mind when interpreting odds ratios.

First, odds ratios tell us about the relative *odds* of seeing Y = 1 (e.g., seeing a person report turning out to vote). This is different than the coefficients from a logistic model which tell us about the *log* of those odds.

Second, we interpret odds ratios in relation to the number 1 rather than 0. Positive effects for X are seen when the odds ratio is *greater* than 1 (that is: an odds ratio \> 1 tells you that it becomes more likely to see Y = 1 when the independent variable increases by one unit). Negative effects for X are seen when the odds ratio is *smaller* than 1 (i.e., Y = 1 becomes less likely to be observed when X increases by one unit). A confidence interval for an odds ratio that includes 1 in its range (as occurs for `gndrMale` and `lrscale` above) indicates a statistically *insignificant* relationship, while an odds ratio that does not include 1 in its range (as occurs for `agea` and `trstplt`) indicates a statistically *significant* relationship.

Third, we interpret odds ratios using multiplication language. For instance, our model indicates that the odds of voting are `r round(tidy(Vote_model_mp, exponentiate = T)[2,2],2)` *times greater* among male respondents than female respondents when holding the effects of age, ideology, and trust constant (although this difference is not statistically significant). Or: the odds of turning out to vote increase by 1.02 times for each one year increase in age (holding constant the other predictor variables).
:::
